---
title: "Exploring Factors Influencing Movie Success"
subtitle: "Stat 184 Final Project"
author: 
  - "Sara Al Riyami"
  - "Layan Al Busaidi"
date: "April 28, 2025"
date-modified: now
date-format: long
format:
  pdf:
    toc: false
    number-depth: 5
    cap-location: top
    fig-align: center
    geometry: 
    - top=1in
    - left=1in
    - right=1in
    - bottom=1in
    colorlinks: true
execute: 
  echo: false
  warning: false
  error: false
---

```{r}
#| label: Load Packages 
#| include: false

# Load all necessary packages -----
library(tidyverse)
library(rvest)
library(dplyr)
library(janitor)
library(knitr)
library(kableExtra)
library(ggplot2)

#Define global elements ----
psuPalette <- c("#1E407C", "#BC204B", "#3EA39E", "#E98300",
                "#999999", "#AC8DCE", "#F2665E", "#99CC00")

```

# Introduction

Movies are a huge part of our lives and a multi-billion-dollar industry that shapes and reflects society. With millions of people watching and reviewing films every year, the movie industry continues to evolve. Thousands of films are released annually, but only a few become truly successful. What factors contribute to a movie's success? Is it the genre, the budget, the actors, or something else? In this project, we set out to explore the elements that influence a movie’s success by analyzing datasets of films and looking at trends in ratings, profits, and casting.

We are interested in this topic because movies are not only a major form of entertainment but also a massive industry. Understanding what leads to success can offer insights into both audience preferences and the strategies that shape film making today.

To guide our analysis, we focused on three key research questions:

1. How do audience ratings compare across the five most common movie genres?

2. Which top studios have the best return on investment, and is there a relationship between movie budget and profit?

3.  Which stars appear most frequently in successful movies, and does their presence correlate with higher success?

Through these questions, we aim to uncover patterns and insights that explain what contributes to a movie’s success. By comparing audience ratings across genres, we hope to identify which types of films tend to get the most viewers. Investigating studio performance and budget-profit relationships allows us to understand how financial decisions impact profitability. Finally, analyzing the presence of frequently featured stars helps us explore whether certain actors consistently appear in higher-performing films, potentially revealing a link between star presence and success. Overall, our goal is to better understand the creative and financial factors that shape successful movies.

# Data Provenance

SARA

## Main Dataset

SARA

## Secondary Dataset

SARA

## Merged and Final Dataset

SARA

```{r}
#| label: LoadingReadingURL

basePart <- "https://raw.githubusercontent.com/Stat184-Spring2025/"
mainPart <- "Sec4_FP_Layan_Sara/main/Data/MoviesJoined.csv"
url <- paste0(basePart,mainPart)
MoviesJoined <- read.csv(url, header = TRUE)

```

# FAIR and CARE Principles

FAIR SARA CARE LAYAN

# EDA: Exploratory Data Analysis

In our project, we used Exploratory Data Analysis to understand the structure of our datasets and uncover any errors or patterns before diving into deeper analysis. We began by cleaning and wrangling the data to ensure it was accurate and usable. This included identifying and removing rows with missing values (NAs), as well as fixing inconsistencies.

After cleaning the data, we created frequency and summary tables to examine key statistics such as the mean, median, and standard deviation, giving us a clearer picture of the overall distribution of variables like ratings. We then used visualizations to uncover patterns and relationships that weren’t immediately apparent from the summary statistics alone.

Through EDA, we were able to detect  irregularities in our dataset, gain a better understanding of its overall structure, and identify key trends that guided our analysis. This process provided a strong foundation for answering our research questions and exploring the factors that contribute to a movie’s success.

## Genre and Rating

LAYAN

### Summary

```{r}
#| label: MakeSummaryTable

# Creating a summary table of Ratings by Genres ----

Genre_summary <- MoviesJoined%>%
  group_by(Genre)%>%      # Groups the data by Genre column
  summarise(           # Calculates summary statistics for each genre
    FilmCount = n(),      # Number of films in each genre
    MinRating = min(Rating, na.rm = TRUE),   #Minimum rating (ignores NA values)
    Q1Rating = quantile(Rating, 0.25, na.rm = TRUE),  # First quartile
    MedianRating = median(Rating, na.rm = TRUE),      # Median rating
    Q3Rating = quantile(Rating, 0.75, na.rm = TRUE),  # Third quartile
    MeanRating = mean(Rating, na.rm = TRUE),     # Mean (average) rating
    MaxRating = max(Rating, na.rm = TRUE),      # Maximum rating
    SdRating = sd(Rating, na.rm = TRUE)        # Standard deviation of ratings
  ) %>%
  arrange(desc(FilmCount))%>%          # Sorts the genres by film count
  slice_head(n=5)       # Selects the top 5 movie genres with the most films

```

```{r}
#| label: tbl-GenreRating
#| tbl-cap: "Summary Table of Ratings for the Top 5 Movie Genres"
#| html-table-processing: none
#| caption-justification: center
#| tbl-pos: "H"

# Displaying the summary table ----
Genre_summary%>%
  kable(
    booktabs = TRUE,                    
    align = c("l", rep("c",8)), # Left-aligns the first column, centers the rest
    format = "latex"
  )%>%
  kableExtra::kable_styling(
    font_size = 16,          # Sets font size of the table
    latex_options = c("striped","scale_down"),
  )%>%
  row_spec(0, bold = TRUE, background = "pink")%>%  # Styles the header
  column_spec(1, italic = TRUE) # Styles the 1 column

```

### Box Plot

```{r}
#| label: WrangleGRBoxPlot

# Wrangling Data ----
## Get Top 5 Genres 
TopGenres <- MoviesJoined %>%
  count(Genre, sort = TRUE) %>%  # Counts num of movies per genre and sorts them
  slice_max(order_by = n, n = 5) %>%  # Selects top 5 genres w most movies 
  pull(Genre)

## Show data for only the Top 5 genres
MovieGenre <- MoviesJoined %>%
  filter(Genre %in% TopGenres)  # Filter movies of only the top 5 genres
  
```

```{r}
#| label: fig-GenreRatingBox
#| fig-cap: "Distribution of Ratings for Top 5 Genres"
#| fig-alt: "Box Plot showing distribution of ratings across top 5 genres"
#| fig-pos: "H"
#| fig-height: 5


# Create the box plot for Genre and Ratings----
ggplot(
  data = MovieGenre,
  mapping = aes(
    x = Rating,      # Set the x-axis to represent Rating
    y = Genre    # Set the y-axis to represent Genre
  )
) +     
geom_boxplot(fill = "lightpink") +   # Creates box plot with pink boxes
labs(                #labels the x and y axis
  y = "Top Genres",
  x = "Rating"
) +
theme_minimal()+
theme(
  text = element_text(size = 12),
  axis.title.x = element_text(face = "bold",   # Make the x-axis title bold
                              size = 14,    # Set font size to 14
                              margin = margin(t = 15)
                              ),
  axis.title.y = element_text(face = "bold", 
                              size = 14, 
                              margin = margin(r = 15) 
                              ) # margin pushes titles away from axis
)

```

## Companies and Movies

SARA

# Q 3

# Conclusion

LAYAN

# Sources and References

LAYAN

{{< pagebreak >}}


# Code Appendix

```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
